Clustering and Parameter Estimation for the Popularity Adjusted Block Model

### 2020-07-19 summary.Rmd
added parameter estimation methods discussed previously, removed lorentz transformation based methods, still working on adding comparisons to other methods

---
title: "Clustering and Parameter Estimation for the Popularity Adjusted Block Model"
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,vlined,ruled]{algorithm2e} 
- \setcitestyle{numbers,square}
bibliography: misc.bib
abstract: |
  In this paper, we connect two probabilistic models for graphs, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  clustering and parameter estimation. In particular, we note that the PABM 
  can be represented as latent positions such that points within the same 
  cluster lie on a subspace, and the subspaces that represent each 
  cluster are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our clustering and 
  parameter estimation methods for the PABM. 
---

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Edge probability matrix $P$, number of clusters $K$}
  \KwResult{Cluster assignments $1, ..., K$}
    Compute the spectral decomposition $P = V D V^\top$.\;
    Compute the inner product matrix $B = |V V^\top|$, applying $|\cdot|$ 
    entry-wise.\;
    Construct graph $G$ using $B$ as edge similarities.\;
    Identify the connected components of $G$ and map each to cluster labels 
    $1, ..., K.$\;
  \caption{PABM clustering on the edge probability matrix}
\end{algorithm}

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of clusters $K$}
  \KwResult{Cluster assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected components 
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the cluster labels $1, ..., K$.\;
  \caption{PABM clustering on the adjacency matrix}
\end{algorithm}

@@

@@

@@

In this part, we will assess Algorithm 1's performance for sample sizes 
$n = 64, 128, 256, 512, 1024$. For each sample size $n$, $100$ sets of 
$\lambda^{(kl)}$'s are drawn and for each set of parameters, one adjacency 
matrix $A$ is drawn and clustered.

@@

@@

```{r q_trans, fig.cap = 'Pair plots of $ZQ$ where $Q$ was found by minimizing objective function (1). This embedding is equivalent to $XU$ up to 90-degree rotations and reflection along axes.'}
XU.prime <- transform.embedding(Z, z, runif(4, -pi, pi))
pairs(XU.prime, col = z, asp = 1)
```

@@

```{r q_from_a, fig.cap = 'Pair plots of $\\hat{Z} $ where $\\hat{Z}$ is the ASE of an adjacency matrix of a PABM and $Q$ is estimated by minimizing objective function (1).'}
XU.hat <- transform.embedding(Zhat, z, runif(4, -pi, pi))
pairs(XU.hat, col = z, asp = 1)

@@

>>>>

---
title: "Clustering and Parameter Estimation for the Popularity Adjusted Block Model"
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,vlined,ruled]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two probabilistic models for graphs, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  clustering and parameter estimation. In particular, we note that the PABM 
  can be represented as latent positions such that points within the same 
  cluster lie on a subspace, and the subspaces that represent each 
  cluster are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our clustering and 
  parameter estimation methods for the PABM. 
---

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Edge probability matrix $P$, number of clusters $K$}
  \KwResult{Cluster assignments $1, ..., K$}
    Compute the spectral decomposition $P = V D V^\top$.\;
    Compute the inner product matrix $B = |V V^\top|$, applying $|\cdot|$ 
    entry-wise.\;
    Construct graph $G$ using $B$ as edge similarities.\;
    Identify the connected components of $G$ and map each to cluster labels 
    $1, ..., K.$\;
  \caption{PABM clustering on the edge probability matrix.}
\end{algorithm}

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of clusters $K$}
  \KwResult{Cluster assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected components 
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the cluster labels $1, ..., K$.\;
  \caption{PABM clustering on the adjacency matrix.}
\end{algorithm}

@@

For any $P$ edge probability matrix for the PABM such that the rows and 
columns are organized by cluster, the $kl$^th^ block is an outer product 
of two vectors, i.e., $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 
Therefore, given $P^{(kl)}$, $\lambda%{(kl)}$ and $\lambda^{(lk)}$ are 
solvable exactly (up to multiplication by $-1$) using singular value 
decomposition. More specifically, let $P = \sigma^2 u v^\top$ be the 
singular value decomposition of $P$. $u \in \mathbb{R}^{n_k}$ and 
$v \in \mathbb{R}^{n_l}$ are vectors and 
$\sigma^2 > 0$ is a scalar. Then $\lambda^{(kl)} = \pm \sigma u$ and 
$\lambda^{(lk)} = \pm \sigma v$.

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the edge probability matrix.}
  \KwData{Edge probability matrix $P$, cluster assignments $1, ..., K$.}
  \KwResult{PABM parameters $\{\lambda^{(kl)}\}_K$}
  Arrange the rows and columns of $P$ by cluster such that each $P^{(kl)}$ 
  block consists of edge probabilities between clusters $k$ and $l$.\; 
  \For{$k, l = 1, ..., K$, $k \geq l$} {
    Compute $P^{(kl)} = (\sigma^{(kl)})^2 u^{(kl)} (v^{(kl)})^\top$, the SVD of 
    the $kl^{th}$ block.\;
    Assign $\lambda^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\lambda^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

A similar method can be applied using $\hat{Z}$, the adjacency spectral 
embedding of $A$.

\begin{algorithm}[H]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation using the adjacency matrix.}
  \KwData{Adjacency matrix $A$, cluster assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(kl)}\}_K$.}
  Assign $p \leftarrow K (K + 1) / 2$ and $q \leftarrow K (K - 1) / 2$.\;
  Compute $\hat{Z}$, the adjacency spectral embedding of $A$ using signature
  $(p, q)$.\;
  Compute $\hat{P} = \hat{Z} I_{pq} \hat{Z}^\top$, the estimate of the edge 
  probability matrix based on the adjacency spectral embedding.\;
  Arrange the rows and columns of $\hat{P}$ by cluster such that each 
  $\hat{P}^{(kl)}$ block consists of estimated edge probabilities between 
  clusters $k$ and $l$.\;
  \For{$k, l = 1, ..., K$, $k \leq l$} {
    Compute $\hat{P}^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
    Assign $\sigma^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

**Theorem 5**. Let $\lambda^{(kl)}_{n_k} \sim F_{n_k}$ for $k, l = 1, ..., K$, 
and let $\{\hat{\lambda}^{(kl)}_{n_k}\}_K$ be the set of estimates computed by 
Algorithm 4. 
Then ... ***TODO***

\begin{comment}

### Using Adjacency Spectral Embedding

@@

\end{comment}

@@

In this part, we will assess Algorithm 2's performance for sample sizes 
$n = 64, 128, 256, 512, 1024$. For each sample size $n$, $100$ sets of 
$\lambda^{(kl)}$'s are drawn and for each set of parameters, one adjacency 
matrix $A$ is drawn and clustered.

@@

\begin{comment}

@@

```{r q_trans, fig.cap = 'Pair plots of $ZQ$ where $Q$ was found by minimizing objective function (1). This embedding is equivalent to $XU$ up to 90-degree rotations and reflection along axes.', cache = TRUE}
XU.prime <- transform.embedding(Z, z, runif(4, -pi, pi))
pairs(XU.prime, col = z, asp = 1)
```

@@

```{r q_from_a, cache = TRUE, fig.cap = 'Pair plots of $\\hat{Z} $ where $\\hat{Z}$ is the ASE of an adjacency matrix of a PABM and $Q$ is estimated by minimizing objective function (1).'}
XU.hat <- transform.embedding(Zhat, z, runif(4, -pi, pi))
pairs(XU.hat, col = z, asp = 1)

@@

\end{comment}

```{r}
estimate.lambda.block <- function(P.block, within = FALSE) {
  if (within) {
    P.svd <- svd(P.block)
    u <- P.svd$u[, 1]
    if (mean(u) < 0) u <- -u
    s <- sqrt(P.svd$d[1])
    return(s * u)
  } else {
    P.svd <- svd(P.block)
    u <- P.svd$u[, 1]
    if (mean(u) < 0) u <- -u
    v <- P.svd$v[, 1]
    if (mean(v) < 0) v <- -v
    s <- sqrt(P.svd$d[1])
    return(list(s * u, s * v))
  }
}

lambda.rmse <- function(lambda.matrix, Phat, clustering) {
  K <- max(clustering)
  n <- nrow(Phat)
  n.vector <- sapply(seq(K), function(k) sum(clustering == k))
  sapply(seq(K), function(k) {
    low.ind.k <- ifelse(k == 1, 1, sum(n.vector[seq(k - 1)]) + 1)
    high.ind.k <- sum(n.vector[seq(k)])
    sapply(seq(k), function(l) {
      P.block <- Phat[clustering == k, clustering == l]
      if (k == l) {
        lambda <- as.numeric(lambda.matrix[seq(low.ind.k, high.ind.k), l])
        lambda.hat <- estimate.lambda.block(P.block, TRUE)
        return(sum((lambda - lambda.hat) ** 2))
      } else {
        low.ind.l <- ifelse(l == 1, 1, sum(n.vector[seq(l = 1)]) + 1)
        high.ind.l <- sum(n.vector[seq(l)])
        lambda.kl <- as.numeric(lambda.matrix[seq(low.ind.k, high.ind.k), l])
        lambda.lk <- as.numeric(lambda.matrix[seq(low.ind.l, high.ind.l), k])
        lambda.hats <- estimate.lambda.block(P.block)
        return(sum((lambda.kl - lambda.hats[[1]]) ** 2) + 
                 sum((lambda.lk - lambda.hats[[2]]) ** 2))
      }
    }) %>% 
      sum() %>% 
      return()
  }) %>% 
    sum() %>% 
    magrittr::divide_by(n * K) %>% 
    sqrt() %>% 
    return()
}
```

Figure \ref{fig:lambda_est} shows the medians and interquartile ranges of root 
mean square errors for Algorithm 4 over 100 simulations using the same 
parameters as before.

```{r lambda_est, cache = TRUE, fig.cap = 'Median and IQR RMSE from Algorithm 4 using sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size.'}
set.seed(314159)

Ipq <- diag(c(1, 1, 1, -1))

rmse.df <- foreach(n = n.vec, .combine = dplyr::bind_rows) %do% {
  plyr::ldply(seq(iter), function(i) {
    z <- sample(c(1, 2), n, prob = c(alpha, 1 - alpha), replace = TRUE)
    z <- sort(z)

    X <- cbind(c(lambda11, rep(0, n2)),
               c(lambda12, rep(0, n2)),
               c(rep(0, n1), lambda21),
               c(rep(0, n1), lambda22))
    Y <- cbind(c(lambda11, rep(0, n2)),
               c(rep(0, n1), lambda21),
               c(lambda12, rep(0, n2)),
               c(rep(0, n1), lambda22))
    P <- X %*% t(Y)
    A <- draw.graph(P)
    lambda.matrix <- cbind(c(lambda11, lambda21),
                           c(lambda12, lambda22))
    Zhat <- embedding(A, p, q)
    Phat <- Zhat %*% Ipq %*% t(Zhat)
    rmse <- lambda.rmse(lambda.matrix, Phat, z)
    dplyr::tibble(n = n, rmse = rmse) %>% 
      return()
  }, .parallel = TRUE) %>% 
    return()
}

rmse.df %>% 
  dplyr::group_by(n) %>% 
  dplyr::summarise(median.rmse = median(rmse),
                   q1.rmse = quantile(rmse, .25),
                   q3.rmse = quantile(rmse, .75)) %>% 
  dplyr::ungroup() %>% 
  ggplot() + 
  geom_line(aes(x = n, y = median.rmse)) + 
  geom_errorbar(aes(x = n, ymin = q1.rmse, ymax = q3.rmse)) + 
  labs(y = 'RMSE')
```
